The goal of this analysis is to identify which chemical properties influence
the quality of red wines.
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.
library(ggplot2)
library(grid)
library(gridExtra)
library(GGally)
library(corrplot)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
redwines<-read.csv('wineQualityReds.csv')
redwines$X <- NULL
```

# Univariate Plots Section
The red wine data set contains 1,599 observations with 11 variables on the
chemical properties of the wine.

As a first step, I will summarise the dataframe, to check the data quality and
if all the variables look reasonable.
Also getting a feeling from the Minimum/Maximum values, Mean and Median.

```{r echo=FALSE, Univariate_Plots}
names(redwines)
summary(redwines)
```

I am going to investigate the quality variable first.

```{r echo=FALSE, Quality_Distribution}
ggplot(aes(x = quality), data = redwines)+
geom_bar(color="black", fill="coral3")+
              scale_x_discrete("Grade",limits=1:10)+
                ylab("Wines") +
                ggtitle("Distribution of Red Wine Quality")
```

It can be noticed that the Quality has a normal distribution
and the median is 6.

Therefore, I created a category of bad (0-4), average (5-6) and good (7-10).

```{r echo=FALSE}
redwines$rating <- ifelse(redwines$quality < 5, 'bad',
                   ifelse(redwines$quality > 6, 'good', 'average'))
redwines$rating <- ordered(redwines$rating,
                          levels = c('bad', 'average', 'good'))
ggplot(aes(x = rating), data = redwines)+
                xlab("Rating Category")+
                geom_bar(colour="black", fill="coral3")+
                ggtitle("Distribution of Red Wine Categories")
```

It can be noticed that the average category presents most of the observations.
This might be an issue as the bad and good samples are limited.

Next, I will plot the distributions of the variables to identify
the carachetistics of the distribution: Normal distribution, Positive or
Negative Skew, as well as understanding if performing operations on the data
could improve the analysis.

```{r echo=FALSE, Chemical_Distributions}
grid.arrange(ggplot(aes(x = fixed.acidity), data = redwines)+
                xlab("Fixed acidity")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.5),
             ggplot(aes(x = volatile.acidity), data = redwines)+
                xlab("volatile.acidity")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.05),
             ggplot(aes(x = citric.acid), data = redwines)+
                xlab("citric.acid")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.05),
             ggplot(aes(x = residual.sugar), data = redwines)+
                xlab("residual.sugar")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.5),
             ggplot(aes(x = chlorides), data = redwines)+
                xlab("chlorides")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.02),
             ggplot(aes(x = free.sulfur.dioxide), data = redwines)+
                xlab("free.sulfur.dioxide")+
                geom_histogram(colour="black", fill="coral3", binwidth = 5),
             ggplot(aes(x = total.sulfur.dioxide), data = redwines)+
                xlab("total.sulfur.dioxide")+
                geom_histogram(colour="black", fill="coral3", binwidth = 20),
             ggplot(aes(x = density), data = redwines)+
                xlab("density")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.001),
             ggplot(aes(x = pH), data = redwines)+
                xlab("pH")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.03),
             ggplot(aes(x = sulphates), data = redwines)+
                xlab("sulphates")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.1),
             ggplot(aes(x = alcohol), data = redwines)+
                xlab("alcohol")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.2),
             ncol=4
             ,top=textGrob("Chemical properties distributions",gp=gpar(fontsize=20)))
```

From all the distribution plots above, I have noticed the following:

Normal Distributions:Volatile acidity, Density, pH.

Positively Skewed: Fixed Acidity, Residual Sugar, Chlorides,
Free Sulphur Dioxide, Total Sulphur Dioxide, Sulphates, Alcohol.

For the moment, I need more investigation on the Citric acid distribution.
I notice a positive skew, however, multiple observations have zero values. 

# Univariate Analysis

For the properties with positive skewed distribution, as they also
contain outliers, I am going to plot the distributions using the
log10 function so to transfor the data to a more normal distribution
and reduce the influence of the outliers.

```{r echo=FALSE, citric.acid, message=FALSE, warning=FALSE}
base_breaks <- function(n = 10){
    function(x) {
        axisTicks(log10(range(x, na.rm = TRUE)), log = TRUE, n = n)
    }
}
grid.arrange(
ggplot(aes(x = citric.acid), data = redwines)+
                xlab("citric.acid")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.05),
ggplot(aes(x = citric.acid), data = redwines)+
                xlab("citric.acid (log)")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.05)
                +scale_x_log10(breaks=base_breaks())
,top=textGrob("Citric acid distribution",gp=gpar(fontsize=20)))
```

As expected, I notice on the plot that Citric Acid contains 132 zero values
and log10(0) is not defined.

```{r, echo=TRUE}
length(which(redwines$citric.acid==0))
```             

This might be due to a wrong data entry, however, many types of red wines
do not contain any citric acid.
The distribution do not appear to be normally distributed with log10 function
applied.

Next, I will proceed with plotting the other distributions.

```{r echo=FALSE}
grid.arrange(
ggplot(aes(x = fixed.acidity), data = redwines)+
                xlab("fixed.acidity")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.5),
ggplot(aes(x = fixed.acidity), data = redwines)+
                xlab("fixed.acidity (log)")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.05)
                +scale_x_log10(breaks=base_breaks())
,top=textGrob("Fixed acidity distribution",gp=gpar(fontsize=20))
,ncol=2)

```

Nothing relevant to note on the Fixed Acidity.

Next, I am going to explore the properties that contain relevant outliers.

```{r echo=FALSE}
grid.arrange(
ggplot(aes(x = free.sulfur.dioxide), data = redwines)+
                xlab("free.sulfur.dioxide")+
                geom_histogram(colour="black", fill="coral3", binwidth = 1),
ggplot(aes(x = free.sulfur.dioxide), data = redwines)+
                xlab("free.sulfur.dioxide (log)")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.05)
                +scale_x_log10(breaks=base_breaks())
,top=textGrob("Free sulfur distribution",gp=gpar(fontsize=20))
,ncol=2)
```

Applying the log10 function, Free sulfur dioxide, resembles to a more normal
distribution.

```{r echo=FALSE}
grid.arrange(
ggplot(aes(x = total.sulfur.dioxide), data = redwines)+
                xlab("total.sulfur.dioxide")+
                geom_histogram(colour="black", fill="coral3", binwidth = 5),
ggplot(aes(x = free.sulfur.dioxide), data = redwines)+
                xlab("total.sulfur.dioxide (log)")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.05)
                +scale_x_log10(breaks=base_breaks())
,top=textGrob("Total sulfur distribution",gp=gpar(fontsize=20))
,ncol=2)
```

Same is for Total sulfur dioxide.

```{r echo=FALSE}
grid.arrange(
ggplot(aes(x = sulphates), data = redwines)+
                xlab("sulphates")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.03),
ggplot(aes(x = sulphates), data = redwines)+
                xlab("sulphates (log)")+
                geom_histogram(colour="black", fill="coral3", binwidth = 0.03)
                +scale_x_log10(breaks=base_breaks())
,top=textGrob("Sulphates distribution",gp=gpar(fontsize=20))
,ncol=2)
```

I do not notice a relevant change in the sulphates distribution, therefore,
in order to get a feeling for the outliers, below is a boxplot distribution.

```{r echo=FALSE}
create_boxplot <- function(column, ylab) {
  return(qplot(data = redwines, x = '',
               y = column, geom = 'boxplot',
               xlab = '',
               ylab = ylab))
}

grid.arrange(create_boxplot(redwines$fixed.acidity, 'fixed acidity'),
             create_boxplot(redwines$volatile.acidity, 'volatile acidity'),
             create_boxplot(redwines$citric.acid, 'citric acid'),
             create_boxplot(redwines$residual.sugar, 'residual sugar'),
             create_boxplot(redwines$chlorides, 'chlorides'),
             create_boxplot(redwines$free.sulfur.dioxide, 'free sulf. dioxide'),
             create_boxplot(redwines$total.sulfur.dioxide, 'total sulf. dioxide'),
             create_boxplot(redwines$density, 'density'),
             create_boxplot(redwines$pH, 'pH'),
             create_boxplot(redwines$sulphates, 'sulphates'),
             create_boxplot(redwines$alcohol, 'alcohol'),
             ncol = 3
             ,top=textGrob("Chemical properties box plots",gp=gpar(fontsize=20)))
```

The boxplots above show that indeed sulphate, do contains outliers, however,
I will investigate on this further on the bivariate analysis.

This plot also helped me in giving less priority on residual sugar and
chlorides, as the distribution do not look interesting for the moment.

Before proceeding in further analysis, the properties that seem to be
more interesting, according to their distributions, are citric acid, fixed and
volatile acidity, pH (as I expect to be correlated with acid) and alcohol, as
it is a commonly known chemical property.

# Bivariate Plots Section

The first step of the bivariate analysis is the correlations to see which ones
seem to be more strongly correlated with quality of course, and to identify
if some expected relationships are confirmed (acids with pH).

```{r echo=FALSE}
correlations = cor(redwines[sapply(redwines, is.numeric)])
corrplot(correlations)
```

From the correlation plot, I can notice already that  quality has
higher correlation on volatile.acidity, alcohol, sulphates, for the other
properties I will compare the actual numbers.

The strongest correlation, by looking at the plot, are
pH and fixed.acidity, fixed.acidity with citric acid.
This suggests me that I will have to explore the acid properties further.

# Bivariate Analysis

Below is the correlation factors with quality, ordered by coefficient. This
first list will help me prioritise the further investigations.

alcohol: 0.47616632

volatile.acidity: -0.390557780

sulphates: 0.251397079

citric.acid: 0.22637251

fixed.acidity: 0.12405165

residual.sugar: 0.013731637

It seems indeed that Alcohol is the most correlated, as well as the volatile
and fixed acidity and Citric Acid.
Surprisingly Sulphates is also correlated. This will require more
investigation.

Below I will plot the correlations with quality, using the category
created to make the plots more readable.
The first plot I believe could be more interesting is the correlation of
Alcohol and Wine Quality.

```{r echo=FALSE}
ggplot(redwines, 
       aes(x=rating, y=alcohol, fill=rating)) + 
       geom_boxplot()+guides(fill=FALSE)+
       xlab("Grade") + ylab("Alcohol (% by volume)") +
       ggtitle("Alcohol and Wine quality correlation")
```

It is indeed quite clear that good quality wines have a higher percentage of
Alcohol.

Next, I will plot the Fixed and Volatile acidity, Citric Acid, pH and Quality
correlation, as the previous analyses showed that they all seem correlated.

```{r echo=FALSE}
grid.arrange(ggplot(data = redwines, aes(x = rating, y = fixed.acidity,
                                   fill = rating)) + 
               ylab('Fixed Acidity (mg / dm^3)') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=FALSE),
             ggplot(data = redwines, aes(x = rating, y = volatile.acidity,
                                   fill = rating)) +
               ylab('Volatile Acidity (mg / dm^3)') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=FALSE), 
             ggplot(data = redwines, aes(x = rating, y = citric.acid,
                                   fill = rating)) +
               ylab('Citric Acid (mg / dm^3)') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=FALSE), 
             ggplot(data = redwines, aes(x = rating, y = pH,
                                   fill = rating)) +
               ylab('pH') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=FALSE),
             top=textGrob("Acidity and pH correlation with Red Wine Quality",gp=gpar(fontsize=20)))
```

It indeed looks interesting to see the positive correlation of
citric acid and Fixed Acidity and a negative correlation of
Volatile acidity. Nothing interesting on the correlation with pH and Quality.

The other chemical property that seems to be correlated is Sulphates. Below
the correlation.
```{r echo=FALSE}
ggplot(redwines, 
       aes(x=rating, y=sulphates, fill=rating)) + 
       geom_boxplot()+guides(fill=FALSE)+
       xlab("Grade") + ylab("sulphates") +
       ggtitle("Sulphates and Wine quality correlation")
```
The plot above shows that good quality red wines contain high sulphates,
this chemical property is used against microbial spoilage.

As said, the most correlated chemical property with quality is alcohol.
I am going to perform a linear model to confirm the hypotesis.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(lm(quality ~ alcohol,redwines))
```

The model above indicates that alcohol explains the 22.6% of the variability
of the quality.

# Multivariate Plots Section

Based on the bivariate findings, I will investigate on integrations between
variables. With more focus on the highly correlated with quality:
alcohol, sulphates, volatile acidity, fixed acidity, citric acid and pH.


```{r echo=FALSE}
create_geom_point_plot <- function(x, y,xlab,ylab,title) {
  return(
          ggplot(data = redwines,
                 aes(x = x, y = y,
                     color = factor(quality))) +
            xlab(xlab) + ylab(ylab)+
            geom_point() +
            facet_wrap(~rating)+
            theme_dark()+
            scale_colour_brewer(palette = "RdYlBu")+
            labs(color='Quality Grade')+
            ggtitle(title)
  )}

create_geom_point_plot(redwines$alcohol, redwines$sulphates,'Alcohol','Sulphates','Alcohol and Sulphate correlation')
```
Higher alcohol and sulphate content are present and positive correlated
in good quality red red wines.

Next is Alcohol and volatile acidity.
```{r echo=FALSE}
create_geom_point_plot(redwines$alcohol, redwines$volatile.acidity,'Alcohol','Volatile Acidity','Alcohol and Volatile Acidity')
```
Nothing particularly interesting in the Volatile Acidity and Alcohol
correlation for the different categories.

```{r echo=FALSE}
create_geom_point_plot(redwines$citric.acid, redwines$volatile.acidity,'Citric Acid','Volatile Acidity','Citric Acid and Volatile Acidity correlation')
```
Citric acid, even though weakly correlated, has an influence with the red wine
quality. However, many red wines do not contain it.

```{r echo=FALSE}
create_geom_point_plot(redwines$pH, redwines$fixed.acidity,'pH','Fixed Acidity','pH and Fixed Acidity')
```

pH does not show an interesting correlation, however good red wines have
this value clustered on lower values.

```{r echo=FALSE}
create_geom_point_plot(redwines$pH, redwines$citric.acid,'pH','Citric Acid','pH and Citric Acid correlation')
```
pH is correlated with Citric Acidi and Fixed Acidity,

```{r echo=FALSE}
create_geom_point_plot(redwines$pH, redwines$volatile.acidity,'pH','Volatile Acidity','pH and Volatile Acidity correlation')
```
pH is not correlated with Volatile Acidity.

# Multivariate Analysis
From this multivariate analysis, one plot that looks interesting is the
Alcohol and Volatile Acidity correlation. Thanks to categorising the rating,
I notice that as there is almost no correlation in the average category,
it seems to be within the good and bad quality wines.

To explore it further, I am goin to zoom in those two categories:
```{r echo=FALSE}
ggplot(data = subset(redwines, rating != 'average'),
       aes(x = volatile.acidity, y = alcohol,
                      color = rating)) +
  geom_point() +
  ggtitle('Alcohol, Volatile Acidity and Wine Quality correlation') +
  xlab('Volatile Acidity') +
  ylab('Alcohol')
```

To confirm my theory, I would like to calculate the correlation coefficient of
volatile acidity and quality on both average and good/bad categories.

```{r echo=FALSE}
average_redwines = subset(redwines, rating == 'average')
cor.test(average_redwines$volatile.acidity, average_redwines$quality,method='pearson')
```
```{r echo=FALSE}
good_and_bad_redwines = subset(redwines, rating != 'average')
cor.test(good_and_bad_redwines$volatile.acidity, good_and_bad_redwines$quality,method='pearson')
```

Average red wines have a correlation of -0.24 and good/bad red wines have
-0.61, which is much stonger.

### Plot One

As the strongest correlation is in Alcohol and Quality, the first Plot
I would like to highlight is the one below
```{r echo=FALSE, Plot_One}
ggplot(redwines, 
       aes(x=rating, y=alcohol, fill=rating)) + 
       geom_boxplot()+guides(fill=FALSE)+
       xlab("Grade") + ylab("Alcohol (% by volume)") +
       ggtitle("Alcohol and Wine quality correlation")
```

### Description One

Good quality wine have higher alcohol level, this might be surprising as there
is no significant difference on alcohol content between average
and bad quality wine.
However, as alcohol only contributes 22.6% of the total quality, more
exploratory data analysis will have to be done on the acidity.

### Plot Two

Data showed that acitidy plays an important role in the red wine quality.
Therefore, below is the second most interesting plot I found
```{r echo=FALSE, Plot_Two}
grid.arrange(ggplot(data = redwines, aes(x = rating, y = fixed.acidity,
                                   fill = rating)) + 
               ylab('Fixed Acidity (mg / dm^3)') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=FALSE),
             ggplot(data = redwines, aes(x = rating, y = volatile.acidity,
                                   fill = rating)) +
               ylab('Volatile Acidity (mg / dm^3)') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=FALSE), 
             ggplot(data = redwines, aes(x = rating, y = citric.acid,
                                   fill = rating)) +
               ylab('Citric Acid (mg / dm^3)') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=FALSE), 
             ggplot(data = redwines, aes(x = rating, y = pH,
                                   fill = rating)) +
               ylab('pH') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=FALSE),
             top=textGrob("Acidity and pH correlation with Red Wine Quality",gp=gpar(fontsize=20)))
```

### Description Two
Volatile Acidity is low on good quality red wines and progressively high in bad
quality red wines.

Critic acid has a higher presence in good quality red wines.

### Plot Three

For my third plot, the focus is on the Volatile Acidity for good and bad red
wines only, as the average category did not show strong correlation with it.

```{r echo=FALSE, Plot_Three}
ggplot(data = subset(redwines, rating != 'average'),
       aes(x = volatile.acidity, y = alcohol,
                      color = rating)) +
  geom_point() +
  ggtitle('Alcohol, Volatile Acidity and Wine Quality correlation') +
  xlab('Volatile Acidity (mg / dm^3)') +
  ylab('Alcohol (% by volume)')
```

### Description Three
From this scatter plot, excluding the average category, we can clearly see
that high quality red wines tend to have higher alcohol content
and a lower volatile acidity.
The main reason for the acidity is that relatively low acidity,
yeast nutrients needed to promote or maintain fermentation,
or a combination of these three characteristics. [1]
Alcohol, on the other hand, adds body. The wine will have a higher viscosity
which is easily seen in watching it bead on the side of the glass.
A high alcohol wine typically tastes fuller bodied than a light-alcohol wine.[2]

------

# Reflection
With the given dataset, I discovered that alcohol, sulphates, and acidity are
the main components that influence the red wine quality, however, it should be
noted that wine quality never comes down to a single factor
and quality is assessed best by tasting by wine experts.

The correlations found seem to be reasonable:

* Alcohol: Ideal alcohol content varies by type of wine and individual taste.
Taste tests have found that the ideal alcohol content averages
around 13.6 percent. [3]

* Volatile acidity: has a negative correlation to wine quality
as it can give an acidic tastes which is often considered a flaw.

* Sulphates: sulphates has a positive correlation with wine quality.
It is often added by winemakers to prevent spoilage and preserve the taste.

* Citic acids: unlike volatile acid, citic acids has positive correlation with
wine quality. It is also used to acidify wines that are too basic
and as a flavor additive.

The main challenge of this analysis is that the number of the observations
were limited to draw more in-depth conclusions. The rating was simplified to
a number from 1-10 and did not have more in-depth information of taste or the
reason of the quality.

Creating a category of bad, average and good, helped to perform clearer
analysis, especially when focussing on bad and good only.

The main information I would like to have for improving my analysis would be
a value of: sweetness, acidity, tanning, fruit and body, so to correlate the
chemical properties with quality further. It would be also interesting to
join the huge dataset of Vivino app, which contains 28 million users who
contribute to rate millions of wines around the globe, as well as providing
prices, which correlating with quality and chemical properties,
could also be a very interesting analysis.

[1]: https://en.wikipedia.org/wiki/Wine
[2]: http://winefolly.com/review/wine-characteristics/
[3]: https://www.theguardian.com/lifeandstyle/2011/jul/05/wine-alcohol-strength-understated
